\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression Models

### 2024-05-22 scratch/notes_john/JSS-article/JSS-article-3.tex
add section on recursive CV to JSS paper

The \code{"modList"} method for \code{cv()} cross-validates several
competing models, not necessarily of the same class, using the same
division of the data into folds. The \code{cv()} function is introduced
in the context of a preliminary example in Section
\ref{preliminary-example-polynomial-regression} of the paper.

Cross-validating mixed-effects models, using the \code{"glmmTMB"},
\code{"lme"}, and \code{"merMod"} methods for \code{cv()}, involves
special considerations that we take up in Section
\ref{cross-validating-mixed-effects-models}.

The \code{"function"} method for \code{cv()}, discussed in Section
\ref{cross-validating-model-specification}, cross-validates a complex
model-specification process that may, for example, involve choice of
data transformations and predictors, or model selection via CV itself.

The \code{"default"} \code{cv()} method works (perhaps with a bit of
coaxing) with many other existing regression-model classes for which
there is an \code{update()} method that accepts a \code{data} argument.
More generally, the \pkg{cv} package is designed to be extensible, as
discussed in Section \ref{extending-the-cv-package}.

@@

The data for the example in this section are drawn from the \pkg{ISLR2}
package for \proglang{R}, associated with \citet{JamesEtAl:2021}. The
presentation here is close (though not identical) to that in the
original source \citep[ Secs. 5.1, 5.3]{JamesEtAl:2021}, and it
demonstrates the use of the \texttt{cv()} function.\footnote{\citet{JamesEtAl:2021}
  use the \texttt{cv.glm()} function in the \pkg{boot} package
  \citep{CantyRipley2022, DavisonHinkley:1997}. Despite its name,
  \texttt{cv.glm()} is an independent function and not a method of a
  \texttt{cv()} generic function.}

@@

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 32663
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.24054
bias-adjusted cross-validation criterion = 19.22691
95% CI for bias-adjusted CV criterion = (15.76529, 22.68854)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
R> boot::cv.glm(Auto, m.auto.glm)$delta
\end{CodeInput}
\begin{CodeOutput}
[1] 19.24821 19.24787
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   hatvalues = cv(m.auto, k="loo"),
+   Woodbury = cv(m.auto, k="loo", method="Woodbury"),
+   naive = cv(m.auto, k="loo", method="naive"),
+   cv.glm = boot::cv.glm(Auto, m.auto.glm),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(hatvalues = cv(m.auto, k = "loo"), :
less accurate nanosecond times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: milliseconds
      expr        min         lq       mean     median         uq        max
 hatvalues   1.015119   1.201546   1.191919   1.206446   1.222333   1.258987
  Woodbury  12.108407  12.390569  19.157808  12.634949  15.289884  73.319275
     naive 219.469269 220.993977 223.441857 223.066916 226.155180 228.677377
    cv.glm 379.740606 383.517075 390.172306 384.790925 386.413397 442.643421
 neval
    10
    10
    10
    10
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Treat CV as analogous to predicting the response for one or more cases
  in a \emph{newly observed cluster}. In this instance, the folds
  comprise one or more whole clusters; we refit the model with all of
  the cases in clusters in the current fold removed; and then we predict
  the response for the cases in clusters in the current fold. These
  predictions are based only on fixed effects because the random effects
  for the omitted clusters are presumably unknown, as they would be for
  data on cases in newly observed clusters.
\item
  Treat CV as analogous to predicting the response for a newly observed
  case in an \emph{existing cluster}. In this instance, the folds
  comprise one or more individual cases, and the predictions can use
  both the fixed and random effects---so-called ``best-linear-unbiased
  predictors'' or ``BLUPs.''
\end{enumerate}

@@

In the process, we merged variables from the school-level and
student-level data sets and created two new school-level variables:
\code{mean.ses}, which is the average SES for students in each school;
and \code{cses}, which is the individual students' SES centered at their
school means. For details, see \citet[Sec. 7.2.2]{FoxWeisberg:2019}.

@@

The \code{"function"} method for \code{cv()} cross-validates a
model-specification process in a general manner. Functions for four such
model-specification processes are included in the package:
\code{selectStepAIC()}, based on the \code{stepAIC()} function in the
\pkg{MASS} package \citep{VenablesRipley:2002}, performs stepwise
predictor selection for regression models; \code{selectTrans()}, based
on the \code{powerTransform()} function in the \pkg{car} package
\citep{FoxWeisberg:2019}, transforms predictors and the response in a
regression model towards normality; \code{selectTransAndStepAIC()}---the
use of which we illustrate in the current section---performs both of
these procedures sequentially; and \code{selectModelList()}---also
illustrated in the current section---uses CV both to select one of
several competing models, and then, recursively, to estimate prediction
error for the selected model. In a vignette on extending the \pkg{cv}
package, we explain how to add model-selection procedures.

@@

\caption[Scatterplot matrix for the numeric variables in the \code{Auto} data]{Scatterplot matrix for the numeric variables in the \code{Auto} data}\label{fig:Auto-explore}
\end{figure}
\end{CodeChunk}

@@

\caption[Component+residual plots for the working model fit to the \code{Auto} data]{Component+residual plots for the working model fit to the \code{Auto} data}\label{fig:Auto-working-model}
\end{figure}
\end{CodeChunk}

@@

We then apply the (rounded) transformations---all, as it turns out,
logs---to the data and re-estimate the model:

@@

The transformed numeric variables are much better-behaved (cf., Figures
\ref{fig:Auto-explore}, on page \pageref{fig:Auto-explore}, and
\ref{fig:Auto-transformed-scatterplot-matrix}):

@@

\begin{itemize}
\tightlist
\item
  \code{data}, the data set to which the model is fit.
\item
  \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number
  generator; as for \code{cv()}, if the seed isn't supplied by the user,
  then a seed is randomly selected and saved.
\item
  Arguments required by the model-selection function: the starting
  \code{working.model} (\code{m.auto}) for transformation and predictor
  selection, the names of the variables---\code{predictors} and
  \code{response}---that are candidates for transformation, and
  \code{AIC=FALSE}, which specifies use of the BIC for model selection.
\end{itemize}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plot(effects::Effect(c("Europe", "political.knowledge"), m.beps,
+             xlevels=list(Europe=1:11, political.knowledge=0:3),
+             fixed.predictors=list(given.values=c(gendermale=0.5))),
+      lines=list(col=c("blue", "red", "orange")),
+      axes=list(x=list(rug=FALSE), y=list(style="stacked")))
\end{CodeInput}
\begin{figure}

@@

As political knowledge increases, voters tend to align their votes more
closely with the party positions on European integration: The
Conservative Party was ``Eurosceptic,'' while Labour and the Liberal
Democrats were supportive of the UK's participation in the EU.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> BayesRule
\end{CodeInput}
\begin{CodeOutput}
function (y, yhat) 
{
    if (!all(y %in% c(0, 1))) 
        stop("response values not all 0 or 1")
    if (any(yhat < 0) || any(yhat > 1)) 
        stop("fitted values outside of interval [0, 1]")
    yhat <- round(yhat)
    result <- mean(y != yhat)
    attr(result, "casewise loss") <- "y != round(yhat)"
    result
}
<bytecode: 0x12000eac8>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}

After doing some error checking, \code{BayesRule()} rounds the predicted
proabability of a 1 (``success'') response in a binary regression model
to 0 or 1 to obtain a categorical prediction and then reports the
proportion of incorrect predictions. Because the Bayes's rule criterion
is an average of casewise components (as, e.g., is the MSE), a
\code{"casewise loss"} attribute is attached to the result, enabling the
computation of bias correction and confidence intervals (as discussion
in Section
\ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).

@@

That is, we simply call the default \texttt{cv()} method with the
\texttt{type} argument properly set. In addition to supplying the
correct \texttt{type} argument, our method sets the default
\texttt{criterion} for the \texttt{cv.multinom()} method to
\texttt{BayesRuleMulti}. Adding the argument \texttt{criterion.name=}
\texttt{deparse(substitute(criterion))} is inessential, but it insures
that printed output will include the name of the criterion function
that's employed, whether it's the default \texttt{BayesRuleMulti} or
something else. Prior to invoking \texttt{NextMethod()}, we called
\texttt{update()} with \texttt{trace=FALSE} to suppress the iteration
history reported by default by \texttt{multinom()}---it would be tedious
to see the iteration history for each fold.

>>>>

\begin{itemize}
\item
  The \code{"lm"} and \code{"glm"} methods are for linear and
  generalized-linear models fit respectively by the standard
  \proglang{R} \code{lm()} and \code{glm()} functions.
\item
  The \code{"modList"} method for \code{cv()} cross-validates several
  competing models, not necessarily of the same class, using the same
  division of the data into folds. The \code{cv()} function is
  introduced in the context of a preliminary example in Section
  \ref{preliminary-example-polynomial-regression} of the paper.
\item
  Cross-validating mixed-effects models, using the \code{"glmmTMB"},
  \code{"lme"}, and \code{"merMod"} methods for \code{cv()}, involves
  special considerations that we take up in Section
  \ref{cross-validating-mixed-effects-models}.
\item
  The \code{"function"} method for \code{cv()}, discussed in Section
  \ref{cross-validating-model-specification}, cross-validates a complex
  model-specification process that may, for example, involve choice of
  data transformations and predictors, or model selection via CV itself.
\item
  The \code{"default"} \code{cv()} method works (perhaps with a bit of
  coaxing) with many other existing regression-model classes for which
  there is an \code{update()} method that accepts a \code{data}
  argument. More generally, the \pkg{cv} package is designed to be
  extensible, as discussed in Section \ref{extending-the-cv-package}.
\end{itemize}

@@

The data for the example in this section are drawn from the \pkg{ISLR2}
package for \proglang{R}, associated with \citet{JamesEtAl:2021}. The
presentation here is close (though not identical) to that in the
original source \citep[ Secs. 5.1, 5.3]{JamesEtAl:2021}, and it
demonstrates the use of the \texttt{cv()} function.\footnote{\citet{JamesEtAl:2021}
  use the \texttt{cv.glm()} function in the \pkg{boot} package
  \citep{CantyRipley2022, DavisonHinkley:1997}. Despite its name,
  \texttt{cv.glm()} is an independent function and not a method of a
  \texttt{cv()} generic function. The \pkg{boot} package is part of the
  standard \proglang{R} distribution.}

@@

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 423102
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.27853
bias-adjusted cross-validation criterion = 19.26291
95% CI for bias-adjusted CV criterion = (15.80408, 22.72175)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
R> boot::cv.glm(Auto, m.auto.glm)$delta # MSE, biased-corrected MSE
\end{CodeInput}
\begin{CodeOutput}
[1] 19.24821 19.24787
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   hatvalues = cv(m.auto, k="loo"),
+   Woodbury = cv(m.auto, k="loo", method="Woodbury"),
+   naive = cv(m.auto, k="loo", method="naive"),
+   cv.glm = boot::cv.glm(Auto, m.auto.glm),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(hatvalues = cv(m.auto, k = "loo"), :
less accurate nanosecond times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: milliseconds
      expr        min         lq       mean     median         uq        max
 hatvalues   1.139062   1.184244   1.202678   1.194248   1.205769   1.297117
  Woodbury  12.106521  12.310701  13.115158  12.337515  12.808810  16.149900
     naive 217.941773 219.491204 225.861284 220.053909 222.015410 278.078974
    cv.glm 373.900074 379.259963 397.907632 382.333036 437.228141 441.112358
 neval
    10
    10
    10
    10
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Treat CV as analogous to predicting the response for one or more cases
  in a \emph{newly observed cluster}. In this instance, the folds
  comprise one or more whole clusters; we refit the model with all of
  the cases in clusters in the current fold removed; and then we predict
  the response for the cases in clusters in the current fold. These
  predictions are based only on fixed effects because the random effects
  for the omitted clusters are construed to be unknown, as they would be
  for data on cases in newly observed clusters.
\item
  Treat CV as analogous to predicting the response for a newly observed
  case in an \emph{existing cluster}. In this instance, the folds
  comprise one or more individual cases, and the predictions can use
  both the fixed and random effects---so-called ``best-linear-unbiased
  predictors'' or ``BLUPs.''
\end{enumerate}

@@

In the process, we merge variables from the school-level and
student-level data sets, and create two new school-level variables:
\code{mean.ses}, which is the average SES for students in each school;
and \code{cses}, which is the individual students' SES centered at their
school means. For details, see \citet[Sec. 7.2.2]{FoxWeisberg:2019}.

@@

The \code{"function"} method for \code{cv()} cross-validates a
model-specification process in a general manner. Functions for four such
model-specification processes are included in the package:
\code{selectStepAIC()}, based on the \code{stepAIC()} function in the
\pkg{MASS} package \citep{VenablesRipley:2002}, performs stepwise
predictor selection for regression models; \code{selectTrans()}, based
on the \code{powerTransform()} function in the \pkg{car} package
\citep{FoxWeisberg:2019}, transforms predictors and the response in a
regression model towards normality; \code{selectTransAndStepAIC()}---the
use of which we illustrate in the current section---performs both of
these procedures sequentially; and \code{selectModelList()}---also
illustrated in the current section---uses CV both to select one of
several competing models, and then, recursively, to estimate prediction
error for the selected model. In a vignette on extending the \pkg{cv}
package, we explain how to specify additional model-selection
procedures.

@@

\caption[Scatterplot matrix for the numeric variables in the \code{Auto} data]{Scatterplot matrix for the numeric variables in the \code{Auto} data. In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression.}\label{fig:Auto-explore}
\end{figure}
\end{CodeChunk}

@@

\caption[Component+residual plots for the working model fit to the \code{Auto} data]{Component+residual plots for the working model fit to the \code{Auto} data. The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.}\label{fig:Auto-working-model}
\end{figure}
\end{CodeChunk}

@@

We then apply the (rounded) transformations---all, as it turns out, logs
(i.e., ``Oth'' powers)---to the data and re-estimate the model:

@@

The transformed numeric variables are much better-behaved (cf., Figure
\ref{fig:Auto-explore}, on page \pageref{fig:Auto-explore}, and Figure
\ref{fig:Auto-transformed-scatterplot-matrix}):

@@

\begin{itemize}
\tightlist
\item
  \code{data}, the data set to which the model is fit.
\item
  \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number
  generator; as for \code{cv()}, if the seed isn't supplied by the user,
  then a seed is randomly selected and saved.
\item
  Arguments required by the model-selection function: the starting
  \code{working.model} (here, \code{m.auto}) for transformation and
  predictor selection; the names of the variables---\code{predictors}
  and \code{response}---that are candidates for transformation; and
  \code{AIC=FALSE}, which specifies use of the BIC for model selection.
\end{itemize}

@@

In Section \ref{comparing-competing-models}, following \citet[Secs. 5.1,
5.3]{JamesEtAl:2021}, we fit polynomial regressions up to degree 10 to
the relationship of \texttt{mpg} to \texttt{horsepower} for the
\texttt{Auto} data, saving the 10 models, named \texttt{m.1} through
\texttt{m.10}, in \texttt{mlist}. We then used \texttt{cv()} to compare
the cross-validated MSE for the 10 models, discovering that the 7th
degree polynomial had the smallest MSE (by a small margin).

If we then select the 7th degree polynomial model, intending to use it
for prediction, the CV estimate of the MSE for this model will be
optimistic. One solution is to cross-validate the process of using CV to
select the ``best'' model---that is, to apply CV to CV recursively. The
function \texttt{selectModelList()}, which is suitable for use with
\texttt{cv()}, implements this idea.

Applying \texttt{selectModelList()} to the \texttt{Auto}
polynomial-regression models, and using 10-fold CV, we obtain:

\begin{CodeChunk}
\begin{CodeInput}
R> recursiveCV.auto <- cv(selectModelList, data=Auto, 
+                        working.model=mlist,
+                        save.model=TRUE, seed=2120)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 2120
\end{CodeOutput}
\begin{CodeInput}
R> recursiveCV.auto
\end{CodeInput}
\begin{CodeOutput}
10-Fold Cross Validation
criterion: mse
cross-validation criterion = 20.01249
bias-adjusted cross-validation criterion = 20.61866
full-sample criterion = 18.74615 
\end{CodeOutput}
\begin{CodeInput}
R> recursiveCV.auto$selected.model
\end{CodeInput}
\begin{CodeOutput}

Coefficients:
         (Intercept)  poly(horsepower, p)1  poly(horsepower, p)2  
              23.446              -120.138                44.090  
poly(horsepower, p)3  poly(horsepower, p)4  poly(horsepower, p)5  
              -3.949                -5.188                13.272  
poly(horsepower, p)6  poly(horsepower, p)7  
              -8.546                 7.981  
\end{CodeOutput}
\begin{CodeInput}
R> cv(mlist[[7]], seed=2120) # same seed for same folds
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 2120
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 18.89822
bias-adjusted cross-validation criterion = 18.85429
full-sample criterion = 18.07817 
\end{CodeOutput}
\end{CodeChunk}

As expected, recursive CV produces a larger estimate of MSE for the
selected 7th degree polynomial model than CV applied directly to this
model.

We can equivalently call \texttt{cv()} with the list of models as its
first argument and set the argument \texttt{recursive=TRUE}:

\begin{CodeChunk}
\begin{CodeInput}
R> cv(mlist, data=Auto, seed=2120, recursive=TRUE, save.model=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 2120
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
cross-validation criterion = 20.01249
bias-adjusted cross-validation criterion = 20.61866
full-sample criterion = 18.74615 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plot(effects::Effect(c("Europe", "political.knowledge"), m.beps,
+             xlevels=list(Europe=1:11, political.knowledge=0:3),
+             fixed.predictors=list(given.values=c(gendermale=0.5))),
+      lines=list(col=c("blue", "red", "orange")), # party colors
+      axes=list(x=list(rug=FALSE), y=list(style="stacked")))
\end{CodeInput}
\begin{figure}

@@

As political knowledge increases, voters tend to align their votes more
closely with the party positions on European integration: The
Conservative Party was relatively Euro-skeptic, while Labour and the
Liberal Democrats were more supportive of the UK's participation in the
EU.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> BayesRule
\end{CodeInput}
\begin{CodeOutput}
function (y, yhat) 
{
    if (!all(y %in% c(0, 1))) 
        stop("response values not all 0 or 1")
    if (any(yhat < 0) || any(yhat > 1)) 
        stop("fitted values outside of interval [0, 1]")
    yhat <- round(yhat)
    result <- mean(y != yhat)
    attr(result, "casewise loss") <- "y != round(yhat)"
    result
}
<bytecode: 0x12e288c00>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}

After doing some error checking, \code{BayesRule()} rounds the predicted
proabability of a 1 (``success'') response in a binary regression model
to 0 or 1 to obtain a categorical prediction, and then reports the
proportion of incorrect predictions. Because the Bayes's rule criterion
is an average of casewise components (as, e.g., is the MSE), a
\code{"casewise loss"} attribute is attached to the result, enabling the
computation of bias correction and confidence intervals (as discussed in
Section
\ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).

@@

That is, we simply call the default \texttt{cv()} method (via
\code{NextMethod()}) with the \texttt{type} argument properly set. In
addition to supplying the correct \texttt{type} argument, our method
sets the default \texttt{criterion} for the \texttt{cv.multinom()}
method to \texttt{BayesRuleMulti}. Adding the argument
\texttt{criterion.name=} \texttt{deparse(substitute(criterion))} is
inessential, but it insures that printed output will include the name of
the criterion function that's employed, whether it's the default
\texttt{BayesRuleMulti} or something else. Prior to invoking
\texttt{NextMethod()}, we called \texttt{update()} with
\texttt{trace=FALSE} to suppress the iteration history reported by
default by \texttt{multinom()}---it would be tedious to see the
iteration history for each fold.
